Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2021-11-22 doc/JFLA2022/paper.tex
elaborate based on final comment by second reviewer

We believe that Coq books can broadly be divided into two categories:
\begin{itemize}
\item \emph{Executable textbooks} that are in effect large, well-commented Coq programs from which other representations of the material (HTML, PDF) are derived using tools.
\item \emph{Traditional textbooks} generated from documentation languages such as LaTeX that are accompanied by standalone Coq formalizations and/or code snippets.
\end{itemize}
Table~\ref{tbl:books} shows a category breakdown for what we believe are the most prominent Coq books in English. Note that several traditional books, such as Coq'Art~\cite{BC04}, used custom tools during their writing process to keep code snippets synchronized with Coq. However, once a book is published, the accompanying code generally takes on a life of its own~\cite{CoqArtCommunity} and may come to substantially diverge from the book.

@@

We believe the \Hydras electronic book has found an attractive set of tradeoffs between the properties of executable books and traditional books. Specifically, in an executable book, the proof assistant language places limits on the structure. For example, Coq may not permit repeating a definition from another file verbatim, since the definition uses variables that are not present in the current context. Thanks to our \alectr based toolchain, the \Hydras book can include ``live'' code fragments at any point in the text. And in contrast to traditional books, we validate the consistency of code fragments during every build of the electronic book. On the one hand, we follow executable books in providing ``continuous delivery'' of new revisions at a high pace. But on the other hand, we also aim to make regular timestamped versions compatible with specific versions of Coq, akin to new editions of traditional books.

Due to our reliance on LaTeX for writing the book, obtaining a book representation in HTML is more cumbersome than for executable books such as Software Foundations~\cite{SF} and Certified Programming with Dependent Types~\cite{chlipalacpdt2011}. Currently, we generate the Coq code documentation in HTML form separately from the book.

>>>>

We believe that Coq books can broadly be divided into two categories:
\begin{itemize}
\item \emph{Executable textbooks} that are in effect large, well-commented Coq programs from which other representations of the material (HTML, PDF) are derived using tools.
\item \emph{Traditional textbooks} generated from documentation languages such as LaTeX that are accompanied by standalone Coq formalizations and/or code snippets.
\end{itemize}
Table~\ref{tbl:books} shows a category breakdown for what we believe are the most prominent Coq books in English. Note that several traditional books, such as Coq'Art, used custom tools during their writing process to keep code snippets synchronized with Coq. However, once a book is published, the accompanying code generally takes on a life of its own~\cite{CoqArtCommunity} and may come to substantially diverge from the book.

@@

We believe the \Hydras electronic book has found an attractive set of tradeoffs between the properties of executable books and traditional books. Specifically, in an executable book, the proof assistant language places limits on the structure. For example, Coq may not permit repeating a definition from another file verbatim, since the definition uses variables that are not present in the current context. Thanks to our \alectr based toolchain, the \Hydras book can include ``live'' code fragments at any point in the text. And in contrast to traditional books, we validate the consistency of code fragments during every build of the electronic book. On the one hand, we follow executable books in providing continuous delivery of new revisions at a high pace. But on the other hand, we also aim to make regular timestamped versions compatible with specific versions of Coq, akin to new editions of traditional books.

The flip side of our approach is that we separate the Coq sources from the LaTeX sources that describe the source code, and our use of snippet delimiters in the Coq source code files may also lower source readability. For books and libraries with small scope and specialized audiences, distributing knowledge representations between sources and documentation formats the way we do in \Hydras may introduce unnecessary overhead without providing additional benefits.
%
More specifically, we currently distribute the book as a PDF generated by LaTeX with links to corresponding HTML documentation of the Coq sources generated by \coqdoc. This means that in contrast to Software Foundations, we cannot easily obtain a uniform single-format representation of all documentation.
